---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# QuasarDeconData

QuasarDeconData bundles processed peripheral blood mononuclear cell (PBMC) deconvolution benchmark datasets together with a combined 10x Genomics single-cell PBMC reference, all formatted for use with QUASAR deconvolution workflows.

Every dataset shares the same six cell-type labels — `Bcells`, `CD4Tcells`, `CD8Tcells`, `Monocytes`, `NK`, and `Unknown` — so the single-cell reference and the bulk benchmarks can be paired directly without relabeling.

## Installation

You can install the development version of QuasarDeconData from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("SergejRuff/QuasarDeconData")
```

## Datasets

| Object | Platform | Features | Samples / Cells | Ground truth | ID element |
|------------|---------------|------------|------------|------------|------------|
| `load_pbmc_sc_ref()` | 10x Genomics scRNA (6k/8k/10k) | 22,629 | 21,978 cells | `celltype` | — |
| `GSE107011` | RNA-seq | 36,128 | 12 | 12 × 6 | `ensembl_ids` |
| `GSE107572` | RNA-seq | 19,423 | 9 | 9 × 6 | — |
| `GSE120502` | RNA-seq | 39,376 | 249 | 249 × 6 | `gene_ids` |
| `GSE65133` | Microarray (Illumina HT-12 V4) | 35,688 | 20 | 20 × 6 | `probe_ids` |

Note: `GSE65133` is microarray intensity data, not counts — keep that in mind when feeding it to count-based deconvolution methods.

## Single-cell reference

The reference is a processed Seurat object combining the 10x 6k, 8k, and 10k healthy-donor PBMC datasets. Cell-type labels live in `celltype` and the dataset of origin in `donor_id`.

```{r ref}
library(QuasarDeconData)

pbmc_sc_ref <- load_pbmc_sc_ref()
pbmc_sc_ref

# cell-type composition
table(pbmc_sc_ref$celltype)

# dataset of origin
table(pbmc_sc_ref$donor_id)
```

## Bulk benchmark datasets

Each bulk dataset is a named list holding `bulk_expression_profiles` (features × samples), `ground_truth_proportions` (samples × cell types), and a gene/probe identifier vector.

```{r bulk-single}
data(GSE107011)

names(GSE107011)

dim(GSE107011$bulk_expression_profiles)
dim(GSE107011$ground_truth_proportions)

# ground-truth proportions sum to 1 per sample
head(GSE107011$ground_truth_proportions)
rowSums(GSE107011$ground_truth_proportions)
```

The remaining datasets follow the same structure:

```{r bulk-all}
data(GSE107572)
data(GSE120502)
data(GSE65133)

# samples x cell types for each benchmark
sapply(
  list(
    GSE107011 = GSE107011,
    GSE107572 = GSE107572,
    GSE120502 = GSE120502,
    GSE65133  = GSE65133
  ),
  function(x) dim(x$ground_truth_proportions)
)
```

All four expose the same cell-type columns (column order varies by dataset):

```{r celltypes}
lapply(
  list(
    GSE107011 = GSE107011,
    GSE107572 = GSE107572,
    GSE120502 = GSE120502,
    GSE65133  = GSE65133
  ),
  function(x) colnames(x$ground_truth_proportions)
)
```